Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fate of human H3N2 influenza virus lineages

### 2018-01-03 paper/manuscript.tex
Move more of the manuscript over to the new LaTex doc

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[11pt]{article}
\usepackage[round,semicolon]{natbib}
\usepackage[margin=1.25in]{geometry}
\usepackage{kpfonts}

\usepackage{seqsplit}
\usepackage{placeins}

\usepackage{newfloat}
\usepackage[labelfont=bf]{caption}
\usepackage{nameref}
\usepackage{rotating}
\usepackage{color}
\usepackage{float}

\setcounter{topnumber}{8}
\setcounter{bottomnumber}{8}
\setcounter{totalnumber}{8}
\renewcommand{\topfraction}{1}
\renewcommand{\bottomfraction}{1}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{1}

\usepackage[font=small,labelfont=bf]{caption}

\usepackage{newfloat}
\DeclareFloatingEnvironment[name={Figure}]{suppfigure}
\renewcommand{\thesuppfigure}{S\arabic{suppfigure}}
\DeclareFloatingEnvironment[name={Table}]{supptable}
\renewcommand{\thesupptable}{S\arabic{supptable}}
\DeclareFloatingEnvironment[name={File}]{suppfile}
\renewcommand{\thesuppfile}{S\arabic{suppfile}}

\definecolor{darkblue}{rgb}{0, 0.0, 0.6}

\usepackage{hyperref}
\hypersetup{colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue}

\usepackage{seqsplit}

\usepackage{array}
\newcolumntype{R}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\usepackage{setspace}

\renewcommand{\topfraction}{1}
\renewcommand{\bottomfraction}{1}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{1}

@@

\maketitle
\onehalfspacing

\begin{abstract}
Abstract text.
\end{abstract}

\section*{INTRODUCTION}
Very rough outline:
\begin{itemize}
\item Mutations are rampant in the evolution of human influenza virus. Seasonal H3N2 influenza virus in particular rapidly accumulates mutations.
\item The evolution of H3N2 is also characterized by clade competition and population turnover.
\item There have been efforts to predict evolutionary success.
\item Mutations that contribute to antigenic evolution largely determine strain success.
\item RNA viruses can accumulate deleterious mutations, and because the influenza does not appreciably recombine, deleterious mutations are linked to beneficial ones.
\item Mutations in HA that impact viral growth may influence strain success.
\item We need to understand the functional impact of mutations in HA.
\item Previously, we measured the effect of all possible single amino-acid mutations to an H1 hemagglutinin from the A/WSN/1933 (H1N1) strain~\citep{thyagarajan2014inherent,doud2016accurate}. 
\item However, this is a highly lab-adapted strain, and the measurements in this strain may not be particularly relevant for studying mutational processes of more contemporaneous strains circulating in the human population.
\item We chose to study the Perth/2009 H3 HA.
\item This also enabled a comparison of how the preferences have shifted for two diverged HA's.
\end{itemize}

\section*{RESULTS}
\label{sec:results}

\subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral replication in cell culture. 
This strain was the H3N2 component of the influenza vaccine from 2010-2012~\comment{add citation}.
Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral replication in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and \comment{supplementary file or dataset with our full reverse genetics plasmid sequence}).
The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.

We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (\comment{supplementary figure with Sanger sequencing}).
We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables the efficient generation of complex influenza virus libraries~\citep{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
These mutant viruses derived all of their non-HA genes from the lab-adapted WSN/1933 strain.
Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which facilitates HA cleavage and activation~\citep{bottcher2006proteolytic, bottcher2010cleavage}.

@@

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figs/dms_overview/dms_overview.pdf}}
\caption{\label{fig:dms_overview}
{\bf Deep mutational scanning experiments of H3 HA.}
(A) We generated Perth/2009 HA mutant virus libraries using a helper virus approach described in~\cite{doud2016accurate}.
We then passaged this pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants. 
Deep sequencing of the variants before and after selection allowed us to infer the H3 site-specific amino-acid preferences.
(B) The DMS experiments were performed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
We also passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to estimate the experimental noise within a biological replicate.
(C) Mutation frequencies of nonsynonymous, stop, and synonymous mutations for the mutant DNA, mutant virus, and unmutated DNA and virus controls. 
There is selection against nonsynonymous and stop codons in the mutant viruses. 
(D) The Pearson correlations for the amino-acid preferences between each pair of replicates are shown. 
The correlation between the technical replicates is comparable to that between biological replicates, indicating that much of the experimental noise comes from passaging at low MOI. 
}
\end{figure}

\subsection*{The HA preferences improve phylogenetic fit of natural H3 sequences}
How well do the Perth/2009 HA preferences inferred from experimental measurements describe the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well experimentally informed codon substitution models (ExpCM's) constructed from our laboratory measurements improve phylogenetic fit of H3N2 evolution~\citep{hilton2017phydms}.
The results in Table~\ref{tab:phydms} show that the ExpCM outperforms conventional substitution models in describing the evolution of human H3N2 HA. 
The ExpCM also optimizes a stringency parameter for the preferences to more closely reflect the strength of selection in nature.
The stringency parameter in the ExpCM is equal to 2.44, which indicates that although the same amino acids are preferred, the strength of selection is more stringent in nature than in our experiments.
Figure~\ref{fig:logoplot} shows a logo plot of the Perth/2009 HA amino-acid preferences rescaled by this stringency parameter.

\begin{table}
\centering
{\small
\begin{tabular}{cccccccc}
\hline
\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Stringency} & \bf{$\omega$} & \bf{$\overline{\omega}$} & \bf{$\omega_{\alpha}$} & \bf{$\omega_{\beta}$} \\ \hline
ExpCM & 0.0 & -8439.33 & $2.44$ & $0.91$ & -- & -- & -- \\
Goldman-Yang M5 & 2166.06 & -9516.36 & -- & -- & $0.36$ & $0.30$ & $0.84$ \\
ExpCM, averaged across sites & 2504.18 & -9691.42 & $0.68$ & $0.32$ & -- & -- & -- \\
Goldman-Yang M0 & 2607.92 & -9738.29 & -- & $0.31$ & -- & -- & -- \\
\hline
\end{tabular}
}
\caption{\label{tab:phydms}
{\bf The site-specific amino-acid preferences are informative for describing human H3N2 evolution in nature.}
We implemented several codon substitution models for phylogenetic fitting of an alignment of human H3N2 HA sequences. 
The maximum likelihood values for each model were compared using the Akaike information criteria ($\Delta$AIC)~\citep{posada2004model}.
An experimentally-informed codon substitution model (ExpCM) built from the preferences averaged across all replicates performs better than conventional substitution models, specifically the M0 and M5 models in~\cite{yang2000codon}.
A non-site-specific ExpCM informed by preferences averaged across all sites performs comparably to the GY94 class of models, indicating the importance of site-specificity in the ExpCM.
The optimized parameters for each model are also shown.
}
\end{table}

A closer examination of the logo plot reveals that the preferences generally agree with existing knowledge about HA's biochemistry.
For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\citep{waterfield1981disulphide} possess high preference for cysteine.
At residues involved in receptor binding, there are strong preferences for the amino acids at sites Y98, D190, W153, and S228~\citep{weis1988structure,martin1998studies}.
A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\citep{kido1992isolation, stech2005new}.
However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
This codon is part of the signal peptide and is cleaved from the mature HA protein.
One possible explanation for why we do not see a strong preference for Met at this site is due to alternative translation initiation occurring at a downstream or upstream start site, as has been described for HA~\citep{girard2011upstream}.
% No upstream or downstream "ATG"'s are present. I did find an in-frame "ATC" one codon upstream of the canonical start site (in the non-coding region), and interestingly, all Perth/2009 HA sequences deposited in GenBank have an "ATT" in this non-coding position, including Seema's Perth HA.
% I6M(HA2) mutant slightly increases the fusion pH (Daniels 1985 Cell)
% We see high preference for T,S,Q over the wildtype Ile at site 226. Although Q226 in an H3 background is characterized as having preference for avian-type receptors, most of these studies were done in older H3 strains such as Aichi. Wu 2017 Cell Host Microbe has shown extensive epistasis in the RBS, specifically in the 220-loop, so perhaps the amino acids tolerated in this loop have shifted over time. Indeed the wildtype amino acids at 225 and 226 are different in Perth compared to HK68 and Aichi.

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}}
\caption{\label{fig:logoplot}
{\bf The site-specific amino-acid preferences of the Perth/2009 HA.}
This logoplot shows the site-specific amino-acid preferences for the averaged replicates rescaled by the stringency parameter (Table 1) estimated by phydms.
The height of each letter is proportional to its preference at that site, and the preferences for all sites are normalized to sum to 1.
The sites are in H3 numbering.
The top overlay bar shows the relative solvent accessibility.
The bottom overlay bar is colored by the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
The letters directly above each logo indicate the wildtype amino acid at that site.
}
\end{figure}

\subsection*{The mutational tolerance of the head and the stalk shows less of a contrast in H3 than in H1}
We next sought to investigate the inherent mutational tolerance of the Perth/2009 HA. 
Figure~\ref{fig:mut_tolerance} shows the mutational tolerance as calculated from the rescaled Perth/2009 H3 preferences and the rescaled WSN/1933 H1 preferences mapped onto the HA crystal structures.
We found antigenic site C and the most distal portion of the globular head near the 190-helix in the Perth/2009 H3 to be tolerant of mutations.
Interestingly, the H3 stalk including the shorter $\alpha$-helix (helix A) is relatively mutationally tolerant compared to the tolerance of the globular head domain. 
This observation suggests that the stalk may be prone to escape from antibodies, and agrees with previous work demonstrating that it is possible to select for antigenic mutants in H3 by broadly-neutralizing stalk-targeting antibodies~\citep{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human}.
% In contrast, the head domain of the WSN/1933 H1 is more mutationally tolerant relative to its stalk domain.

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figs/mut_tolerance/entropy_heatmap.pdf}}
\caption{\label{fig:mut_tolerance}
{\bf Mutational tolerance mapped onto each site of HA.}
Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}) and the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}), with both trimers in approximately the same orientation. 
The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) from this study, or the preferences measured in the WSN/1933 H1 (right panel) from~\cite{doud2016accurate}. 
Lighter shades of blue or red signify low mutational tolerance, while darker shades of blue or red signify high mutational tolerance. 
For each HA, the structure on the left side colors the full HA trimer, while the structure on the right side colors only one of the monomers.
The sialic acid receptor is shown as black sticks.
The Perth/2009 H3 shows relatively high mutational tolerance in the stalk region, particularly in helix A, compared to the head region. 
High mutational tolerance in H3 was also observed near antigenic site C and near the 190 helix.
The head region of the WSN/1933 H1 is mutationally tolerant compared to the relatively intolerant stalk region. 
}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.75\textwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}}
\caption{\label{fig:trunkvssidebranch}
{\bf Mutations in evolutionary successful strains tend to be more favorable than in strains that die out.}
(A) Phylogenetic tree of human H3N2 influenza virus from 1968-present. 
The trunk is shown in red, and side branches are shown in blue.
The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
(B) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
Negative numbers signify mutations towards less preferred amino acids, while positive numbers signify more preferred mutations.
The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
(C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
The preferences were randomized 10,000 times to estimate significance.
The effects of trunk mutations are higher than side branch internal and terminal node mutations.
}
\end{figure}

How distantly can the preferences be extended to describe differences in successful and unsuccessful strains?
To explore this question, we scored the complete HA sequence of every node in the tree using the preferences by quantifying a \textit{sequence preference} metric.
The sequence preference is defined as 
$$\displaystyle\sum_{r} \ln \pi_{r, a}$$
where $\pi_{r, a}$ is the preference for amino acid $a$ at site $r$.
Consistent with the finding that trunk mutations are generally more favorable than side branch mutations, Figure~\ref{fig:sequence_preference} shows that the sequences of trunk nodes also tend to be more highly preferred than those of side branch nodes.
Interestingly, the sequence preferences increase over time as the nodes approach the Perth/2009 strain and its closely related nodes, which all exhibit sequence preferences that are higher than that of the trunk. 
The observation that the strain in which we performed our deep mutational scan has one of the highest sequence preferences illustrates epistatic interactions among mutations such that an unpreferred mutation in one background may be preferred in another.
% Due to the sheer number of possible pairwise combinations of mutations, it is impossible for us to experimentally characterize these epistatic interactions among mutations.

@@

\begin{figure}
\centerline{\includegraphics[width=0.75\textwidth]{figs/sequence_preference/sequence_preference.pdf}}
\caption{\label{fig:sequence_preference}
{\bf The HA sequences of trunk nodes tend to be more preferred than those of side branch nodes.}
We used the preferences to calculate the HA sequence preference normalized by the number of sites of every node in a human H3N2 phylogenetic tree.
To calculate the sequence preference, we used the entire HA sequence (top panel), only epitope sites (center panel) or non-epitope sites (bottom panel) as defined by~\cite{wolf2006long}.
Higher preferences are closer to zero while lower preferences are more negative.
The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain, indicating epistasis among epitope sites.
The trunk nodes generally exhibit higher sequence preferences than do side branch nodes.
}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=\textwidth]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}}
\caption{\label{fig:WSN_trunkvssidebranch}
{\bf The WSN/1933 H1 preferences do not reveal differences in trunk vs side branch mutational effects}
(A) We used the WSN/1933 H1 preferences from~\cite{doud2016accurate} to calculate the log$_{2}$ mutational effects of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch}.
There is not a significant difference in trunk vs side branch mutational effects.
(B) We also performed a sliding window analysis with the WSN/1933 H1 preferences.
There is not a distinct difference in trunk and side branch mutational effects.
}
\end{figure}

\subsection*{The H1 and H3 preferences at many sites have shifted}
How shifted are the preferences between evolutionarily distant homologs such as H1 and H3?
Although we have previously compared the preferences between related protein homologs of NP~\citep{doud2015site} and of Env \comment{cite haddox2017}, the HA homologs we have experimentally studied are considerably more diverged than either of these pairs.
The WSN/1933 H1 and the Perth/2009 H3 fall into two phylogenetically distinct groups and share 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A), providing an opportunity to analyze what has shifted and what has remained conserved.
Simply correlating the preferences between H1 and H3 reveals that the replicate measurements within a homolog are more correlated than between homologs (Figure~\ref{fig:distance_distribution}B).

>>>>

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template 
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

@@

@@

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{T}his PNAS journal template is provided to help you write your work in the correct journal format.  Instructions for use are provided below.

Note: please start your introduction without including the word ``Introduction'' as a section heading (except for math articles in the Physical Sciences section); this heading is implied in the first paragraphs.

\section*{Results}
\label{sec:results}

\subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral replication in cell culture. 
This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral replication in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and \comment{supplementary file or dataset with our full reverse genetics plasmid sequence}).
The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.

We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (\comment{supplementary figure with Sanger sequencing}).
We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables the efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
These mutant viruses derived all of their non-HA genes from the lab-adapted WSN/1933 strain.
Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which facilitates HA cleavage and activation~\cite{bottcher2006proteolytic, bottcher2010cleavage}.

@@

\begin{figure*}
\centering
\includegraphics[width=13cm]{figs/dms_overview/dms_overview.pdf}
\caption{\label{fig:dms_overview}
{\bf Deep mutational scanning experiments of H3 HA.}
(A) We generated Perth/2009 HA mutant virus libraries using a helper virus approach described in~\cite{doud2016accurate}.
We then passaged this pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants. 
Deep sequencing of the variants before and after selection allowed us to infer the H3 site-specific amino-acid preferences.
(B) The DMS experiments were performed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
We also passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to estimate the experimental noise within a biological replicate.
(C) Mutation frequencies of nonsynonymous, stop, and synonymous mutations for the mutant DNA, mutant virus, and unmutated DNA and virus controls. 
There is selection against nonsynonymous and stop codons in the mutant viruses. 
(D) The Pearson correlations for the amino-acid preferences between each pair of replicates are shown. 
The correlation between the technical replicates is comparable to that between biological replicates, indicating that much of the experimental noise comes from passaging at low MOI. 
}
\end{figure*}

@@

\subsection*{The HA preferences improve phylogenetic fit of natural H3 sequences}
How well do the Perth/2009 HA preferences inferred from experimental measurements describe the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well experimentally informed codon substitution models (ExpCM's) constructed from our laboratory measurements improve phylogenetic fit of H3N2 evolution~\cite{hilton2017phydms}.
The results in Table~\ref{tab:phydms} show that the ExpCM outperforms conventional substitution models in describing the evolution of human H3N2 HA. 
The ExpCM also optimizes a stringency parameter for the preferences to more closely reflect the strength of selection in nature.
The stringency parameter in the ExpCM is equal to 2.44, which indicates that although the same amino acids are preferred, the strength of selection is more stringent in nature than in our experiments.
Figure~\ref{fig:logoplot} shows a logo plot of the Perth/2009 HA amino-acid preferences rescaled by this stringency parameter.

\begin{figure*}
\centering
\caption{\label{tab:phydms}
{\bf The site-specific amino-acid preferences are informative for describing human H3N2 evolution in nature.}}
\begin{tabular}{cccccccc}
\hline
\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Stringency} & \bf{$\omega$} & \bf{$\overline{\omega}$} & \bf{$\omega_{\alpha}$} & \bf{$\omega_{\beta}$} \\ \hline
ExpCM & 0.0 & -8439.33 & $2.44$ & $0.91$ & -- & -- & -- \\
Goldman-Yang M5 & 2166.06 & -9516.36 & -- & -- & $0.36$ & $0.30$ & $0.84$ \\
ExpCM, averaged across sites & 2504.18 & -9691.42 & $0.68$ & $0.32$ & -- & -- & -- \\
Goldman-Yang M0 & 2607.92 & -9738.29 & -- & $0.31$ & -- & -- & -- \\
\hline
\end{tabular}

\addtabletext{We implemented several codon substitution models for phylogenetic fitting of an alignment of human H3N2 HA sequences. 
The maximum likelihood values for each model were compared using the Akaike information criteria ($\Delta$AIC)~\cite{posada2004model}.
An experimentally-informed codon substitution model (ExpCM) built from the preferences averaged across all replicates performs better than conventional substitution models, specifically the M0 and M5 models in~\cite{yang2000codon}.
A non-site-specific ExpCM informed by preferences averaged across all sites performs comparably to the GY94 class of models, indicating the importance of site-specificity in the ExpCM.
The optimized parameters for each model are also shown.}
\end{figure*}

A closer examination of the logo plot reveals that the preferences generally agree with existing knowledge about HA's biochemistry.
For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
At residues involved in receptor binding, there are strong preferences for the amino acids at sites Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies}.
A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\cite{kido1992isolation, stech2005new}.
However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
This codon is part of the signal peptide and is cleaved from the mature HA protein.
One possible explanation for why we do not see a strong preference for Met at this site is due to alternative translation initiation occurring at a downstream or upstream start site, as has been described for HA~\cite{girard2011upstream}.
% No upstream or downstream "ATG"'s are present. I did find an in-frame "ATC" one codon upstream of the canonical start site (in the non-coding region), and interestingly, all Perth/2009 HA sequences deposited in GenBank have an "ATT" in this non-coding position, including Seema's Perth HA.
% I6M(HA2) mutant slightly increases the fusion pH (Daniels 1985 Cell)
% We see high preference for T,S,Q over the wildtype Ile at site 226. Although Q226 in an H3 background is characterized as having preference for avian-type receptors, most of these studies were done in older H3 strains such as Aichi. Wu 2017 Cell Host Microbe has shown extensive epistasis in the RBS, specifically in the 220-loop, so perhaps the amino acids tolerated in this loop have shifted over time. Indeed the wildtype amino acids at 225 and 226 are different in Perth compared to HK68 and Aichi.

\begin{figure*}
\centering
\includegraphics[width=17cm]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}
\caption{\label{fig:logoplot}
{\bf The site-specific amino-acid preferences of the Perth/2009 HA.}
This logoplot shows the site-specific amino-acid preferences for the averaged replicates rescaled by the stringency parameter (Table 1) estimated by phydms.
The height of each letter is proportional to its preference at that site, and the preferences for all sites are normalized to sum to 1.
The sites are in H3 numbering.
The top overlay bar shows the relative solvent accessibility.
The bottom overlay bar is colored by the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
The letters directly above each logo indicate the wildtype amino acid at that site.
}
\end{figure*}

\subsection*{The mutational tolerance of the head and the stalk shows less of a contrast in H3 than in H1}
We next sought to investigate the inherent mutational tolerance of the Perth/2009 HA. 
Figure~\ref{fig:mut_tolerance} shows the mutational tolerance as calculated from the rescaled Perth/2009 H3 preferences and the rescaled WSN/1933 H1 preferences mapped onto the HA crystal structures.
We found antigenic site C and the most distal portion of the globular head near the 190-helix in the Perth/2009 H3 to be tolerant of mutations.
Interestingly, the H3 stalk including the shorter $\alpha$-helix (helix A) is relatively mutationally tolerant compared to the tolerance of the globular head domain. 
This observation suggests that the stalk may be prone to escape from antibodies, and agrees with previous work demonstrating that it is possible to select for antigenic mutants in H3 by broadly-neutralizing stalk-targeting antibodies~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human}.
% In contrast, the head domain of the WSN/1933 H1 is more mutationally tolerant relative to its stalk domain.

@@

\begin{figure*}
\centering
\includegraphics[width=16cm]{figs/mut_tolerance/entropy_heatmap.pdf}
\caption{\label{fig:mut_tolerance}
{\bf Mutational tolerance mapped onto each site of HA.}
Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}) and the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}), with both trimers in approximately the same orientation. 
The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) from this study, or the preferences measured in the WSN/1933 H1 (right panel) from~\cite{doud2016accurate}. 
Lighter shades of blue or red signify low mutational tolerance, while darker shades of blue or red signify high mutational tolerance. 
For each HA, the structure on the left side colors the full HA trimer, while the structure on the right side colors only one of the monomers.
The sialic acid receptor is shown as black sticks.
The Perth/2009 H3 shows relatively high mutational tolerance in the stalk region, particularly in helix A, compared to the head region. 
High mutational tolerance in H3 was also observed near antigenic site C and near the 190 helix.
The head region of the WSN/1933 H1 is mutationally tolerant compared to the relatively intolerant stalk region. 
}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=12cm]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
\caption{\label{fig:trunkvssidebranch}
{\bf Mutations in evolutionary successful strains tend to be more favorable than in strains that die out.}
(A) Phylogenetic tree of human H3N2 influenza virus from 1968-present. 
The trunk is shown in red, and side branches are shown in blue.
The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
(B) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
Negative numbers signify mutations towards less preferred amino acids, while positive numbers signify more preferred mutations.
The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
(C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
The preferences were randomized 10,000 times to estimate significance.
The effects of trunk mutations are higher than side branch internal and terminal node mutations.
}
\end{figure*}

How distantly can the preferences be extended to describe differences in successful and unsuccessful strains?
To explore this question, we scored the complete HA sequence of every node in the tree using the preferences by quantifying a \textit{sequence preference} metric.
The sequence preference is defined as 
$\displaystyle\sum_{r} \ln \pi_{r, a}$
where $\pi_{r, a}$ is the preference for amino acid $a$ at site $r$.
Consistent with the finding that trunk mutations are generally more favorable than side branch mutations, Figure~\ref{fig:sequence_preference} shows that the sequences of trunk nodes also tend to be more highly preferred than those of side branch nodes.
Interestingly, the sequence preferences increase over time as the nodes approach the Perth/2009 strain and its closely related nodes, which all exhibit sequence preferences that are higher than that of the trunk. 
The observation that the strain in which we performed our deep mutational scan has one of the highest sequence preferences illustrates epistatic interactions among mutations such that an unpreferred mutation in one background may be preferred in another.
% Due to the sheer number of possible pairwise combinations of mutations, it is impossible for us to experimentally characterize these epistatic interactions among mutations.

@@

\begin{figure*}
\centering
\includegraphics[width=12cm]{figs/sequence_preference/sequence_preference.pdf}
\caption{\label{fig:sequence_preference}
{\bf The HA sequences of trunk nodes tend to be more preferred than those of side branch nodes.}
We used the preferences to calculate the HA sequence preference normalized by the number of sites of every node in a human H3N2 phylogenetic tree.
To calculate the sequence preference, we used the entire HA sequence (top panel), only epitope sites (center panel) or non-epitope sites (bottom panel) as defined by~\cite{wolf2006long}.
Higher preferences are closer to zero while lower preferences are more negative.
The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain, indicating epistasis among epitope sites.
The trunk nodes generally exhibit higher sequence preferences than do side branch nodes.
}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=15cm]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}
\caption{\label{fig:WSN_trunkvssidebranch}
{\bf The WSN/1933 H1 preferences do not reveal differences in trunk vs side branch mutational effects}
(A) We used the WSN/1933 H1 preferences from~\cite{doud2016accurate} to calculate the log$_{2}$ mutational effects of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch}.
There is not a significant difference in trunk vs side branch mutational effects.
(B) We also performed a sliding window analysis with the WSN/1933 H1 preferences.
There is not a distinct difference in trunk and side branch mutational effects.
}
\end{figure*}

\subsection*{The H1 and H3 preferences have shifted at many sites}
How shifted are the preferences between evolutionarily distant homologs such as H1 and H3?
Although we have previously compared the preferences between related protein homologs of NP~\citep{doud2015site} and of Env \comment{cite haddox2017}, the HA homologs we have experimentally studied are considerably more diverged than either of these pairs.
The WSN/1933 H1 and the Perth/2009 H3 fall into two phylogenetically distinct groups and share 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A), providing an opportunity to analyze what has shifted and what has remained conserved.
Simply correlating the preferences between H1 and H3 reveals that the replicate measurements within a homolog are more correlated than between homologs (Figure~\ref{fig:distance_distribution}B).
